---
# title:
output:
  bookdown::pdf_document2:
    toc: false
    includes:
      in_header: "preamble.tex"
    latex_engine: lualatex
  word_document:
    reference_docx: style_reference.docx
always_allow_html: true
bibliography: bibliography.bib
csl: nlm.csl
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, results = "hide", message = FALSE, warning = FALSE
)
```

```{r source}
library(here)
library(bookdown)
source(here("code", "02_explore.R"))
```

<!--
<center>
Lucas Veras$^{1,2}$, Florêncio Diniz-Sousa$^{1,2}$, Giorjines Boppre$^{1,2}$, Vítor Devezas$^3$, Hugo Santos-Sousa$^3$, John Preto$^3$, João Paulo Vilas-Boas$^{4,5}$, Leandro Machado$^{4,5}$, José Oliveira$^{1,2}$, Hélder Fonseca$^{1,2}$
</center>

\bigskip

1 - Research Center in Physical Activity, Health and Leisure (CIAFEL), Faculty of Sport, University of Porto, Porto, Portugal

2 - Laboratory for Integrative and Translational Research in Population Health (ITR), University of Porto, Porto, Portugal

3 - General Surgery Department, São João Medical Center, Porto, Portugal

4 - Center of Research, Education, Innovation and Intervention in Sport (CIFI2D), Faculty of Sport, University of Porto, Porto, Portugal

5 - Biomechanics Laboratory (LABIOMEP-UP), University of Porto, Porto, Portugal

\leftskip=0.5in \rightskip=0.5in

\bigskip
-->


# Methods {.unnumbered}

A total of `r max(sample_size$pGRF_N)` adults were enrolled in the study (`r unname(sample_size_sex["M"])` males; `r sample_descriptives$age_mean`±`r sample_descriptives$age_sd` years; `r sample_descriptives$height_mean`±`r sample_descriptives$height_sd` cm; `r sample_descriptives$body_mass_mean`±`r sample_descriptives$body_mass_sd` kg; `r sample_descriptives$BMI_mean`±`r sample_descriptives$BMI_sd` kg·m^-2^, X̄±SD). The protocol was approved by the local Ethics Committee (CES 192-14), and all participants were informed about the experiments purpose and protocol before giving written informed consent. No participant reported any musculoskeletal limitations. The measurement of height (stadiometer model 213; Seca, Hamburg, Germany) and body mass (digital scale model 899; Seca, Hamburg, Germany) was accomplished according to standard practice.

After familiarization, participants performed a series of trials envolving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed from the floor to 5cm, 15cm and 30cm high boxes. Finally, continuous jumps were carried out in two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps and both phases of the continuous jumps were recorded in force plates (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions to arms movement, land two-footed and then remain stationary on the force plate for five seconds (except in the continuous jumps).

Throughout the protocol, three activity monitors (GT9X Link; 100Hz; ±16g range; ActiGraph, Pensacola, USA) were worn by each participant, at the following placements: i) at their right hip (along the anterior axillary line, at the level of the iliac crest); ii) at their lower back (at the midpoint between the two posterior superior iliac spines); and iii) at their right ankle (immediately superior to the lateral malleolus). Monitors worn at hip and lower back were positioned on the same elastic belt with clips, and the ankle-worn monitor was fixed by an elastic belt and adhesive tape. Activity monitors were always positioned to ensure the alignment of the accelerometer vertical axis and the standing body craniocaudal axis.

The activity monitors used include a primary and a secondary triaxial accelerometers. As the manufacturer proprietary filter is not applied on the secondary accelerometer raw data, only data from this accelerometer were used in this study, to improve data processing replicability. Data from FP and accelerometers were operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA) and exported as raw data from the *x*, *y* and *z* vectors. GRF was expressed in N and acceleration (ACC) in gravitational ACC units (1*g* = 9.807 m·s^-2^).

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the GRF signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the GRF and ACC resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak ACC (pACC) were determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Peak GRF (pGRF) were then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The peak loading rate (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

\[f_i^{'} = \frac{f_{i + 1} - f_{i - 1}}{t_{i + 1} - t_{i - 1}}\]

\[f_{i_{max}}^{'} = max(f_i^{'})\]

Finally, pGRF, pACC, pLR and pAR means of the resultant and its vertical component for each participant at each jump type and height for each accelerometer placement were extracted and used in all remaining analyses.

Statistical analyses were performed through the R statistical software (version `r paste(R.Version()$major, R.Version()$minor, sep = ".")`, R Foundation for Statistical Computing, Vienna, Austria). All code used to generate the results can be found in an open platform (https://github.com/verasls/jumping_GRF_ACC)(ref). Statistical significance was set as $\alpha$ = 0.05.

Linear mixed models were used to develop pGRF and pLR prediction equations for each vector (resultant and vertical) and accelerometer placement (ankle, lower back and  hip). Body mass was tested as predictor in all models, along with the acceleration magnitude (pACC) in the pGRF models and its rate (pAR) in the pLR models. These predictors were entered as fixed effects and kept in the model, as all have shown to be significant (*p* < 0.05). Subject and the interaction between jump type and height were tested as random effects, and both factors have shown to improve the models. Final models were chosen according to the -2 log-likelihood statistic [@Field2012]. The conditional coefficient of determination (R^2^), which estimates the variance explained by both fixed and random factors [@Nakagawa2013], was computed.

Model validation was performed using the leave-one-out cross-validation [@Staudenmayer2012]. This method consists in separating data from one of the subjects into a testing dataset, and keeping the remainder in the training dataset. Then, a new model is developed using the training dataset, with the same parameters that were defined using the whole sample. This new model is then applied to predict the outcome for the subject in the testing dataset. The whole process is repeated once for each participant (`r max(sample_size$pGRF_N)` times). Data from the testing dataset were used in all subsequent analyses.

The quality of the models was then evaluated both numerically and graphically. The numerical assessment was made by computing the following accuracy indices: the mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE). To visualize the agreement between the values predicted by the models and measured by the FP, Bland-Altman plots were drawn. This type of plot is made by plotting the mean of the actual and predicted values on the x axis and their difference on the y axis. Then, the bias was calculated as the mean of these differences and plotted as a continuous horizontal line, and the limits of agreement were defined by bias ± 1.96 standard deviation and plotted as horizontal dashed lines [@Bland1986]. To further test if there were any consistent under- or over-estimation of the values by the models, one-sample *t*-tests were done to check whether the bias were different than zero. Finally, linear regressions were employed to examine if the difference between actual and predicted values were affected by their magnitude, verifying, therefore, whether the prediction was constant throughout the magnitude range that was tested [@Giavarina2015].

\newpage

# References {.unnumbered}

<div id="refs"></div>
