---
title: "Using raw accelerometer data to predict high-impact mechanical loading"
output:
  bookdown::pdf_document2:
    toc: false
    includes:
      in_header: "preamble.tex"
    latex_engine: lualatex
  word_document:
    reference_docx: style_reference.docx
always_allow_html: true
bibliography: bibliography.bib
csl: nlm.csl
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, results = "hide", message = FALSE, warning = FALSE
)
```

```{r source}
library(here)
library(bookdown)
library(broman)
source(here("code", "02_explore.R"))
```

```{r GRF_accuracy}
load(here("output", "prediction_models.rda"))
res_GRF_accuracy <- purrr::map(
  accuracy_res_GRF_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("ankle", "lower back", "hip"),
    vector = "resultant"
  )
ver_GRF_accuracy <- purrr::map(
  accuracy_ver_GRF_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("ankle", "lower back", "hip"),
    vector = "vertical"
  )
GRF_accuracy <- rbind(res_GRF_accuracy, ver_GRF_accuracy)
hip_GRF_accuracy <- GRF_accuracy %>%
  filter(placement == "hip" & vector == "resultant")
```

```{r LR_accuracy}
load(here("output", "prediction_models.rda"))
res_LR_accuracy <- purrr::map(
  accuracy_res_LR_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("ankle", "lower back", "hip"),
    vector = "resultant"
  )
ver_LR_accuracy <- purrr::map(
  accuracy_ver_LR_models,
  ~ dplyr::mutate(.x, MAPE = as.numeric(MAPE))
) %>%
  purrr::map_dfr(rbind) %>%
  dplyr::mutate(
    placement = c("ankle", "lower back", "hip"),
    vector = "vertical"
  )
LR_accuracy <- rbind(res_LR_accuracy, ver_LR_accuracy)
hip_LR_accuracy <- LR_accuracy %>%
  filter(placement == "hip" & vector == "resultant")
```

```{r ba_agreement}
load(here("output", "loocv_data.rda"))
percent_inside_loa <- function(data) {
  lower <- lvmisc::loa(
    data[[".actual"]], data[[".predicted"]], na.rm = TRUE
  )$lower
  upper <- lvmisc::loa(
    data[[".actual"]], data[[".predicted"]], na.rm = TRUE
  )$upper
  diff <- dplyr::transmute(
    data, diff = .data[[".actual"]] - .data[[".predicted"]]
  )
  diff <- dplyr::mutate(
    diff,
    inside = as.factor(ifelse(diff > lower & diff < upper, "yes", "no"))
  )
  round(prop.table(table(diff$inside))[[2]] * 100, 0)
}

ba_agreement <- min(
  map_dbl(cv_res_GRF_models, percent_inside_loa) %>% min(),
  map_dbl(cv_ver_GRF_models, percent_inside_loa) %>% min(),
  map_dbl(cv_res_LR_models, percent_inside_loa) %>% min(),
  map_dbl(cv_ver_LR_models, percent_inside_loa) %>% min()
)
```

```{r ba_regression}
source(here("code", "04_analyse_results.R"))
ba_regression_ankle_R2 <- c(
  ba_regression_res_GRF$ankle["adj.r.squared"],
  ba_regression_ver_GRF$ankle["adj.r.squared"],
  ba_regression_res_LR$ankle["adj.r.squared"],
  ba_regression_ver_LR$ankle["adj.r.squared"]
) %>%
  as_vector()
```

<!--
<center>
Lucas Veras$^{1,2}$, Florêncio Diniz-Sousa$^{1,2}$, Giorjines Boppre$^{1,2}$, Vítor Devezas$^3$, Hugo Santos-Sousa$^3$, John Preto$^3$, João Paulo Vilas-Boas$^{4,5}$, Leandro Machado$^{4,5}$, José Oliveira$^{1,2}$, Hélder Fonseca$^{1,2}$
</center>

\bigskip

1 - Research Center in Physical Activity, Health and Leisure (CIAFEL), Faculty of Sport, University of Porto, Porto, Portugal

2 - Laboratory for Integrative and Translational Research in Population Health (ITR), University of Porto, Porto, Portugal

3 - General Surgery Department, São João Medical Center, Porto, Portugal

4 - Center of Research, Education, Innovation and Intervention in Sport (CIFI2D), Faculty of Sport, University of Porto, Porto, Portugal

5 - Biomechanics Laboratory (LABIOMEP-UP), University of Porto, Porto, Portugal

\leftskip=0.5in \rightskip=0.5in

\bigskip
-->

# Introduction {.unnumbered}

Bone responds to mechanical loading by altering its properties, such as mass, microarchitecture, and geometry, which ultimately dictate its strength and resistance to fracture [@Turner2003; @Fonseca2014; @Santos2017]. In fact, physical activity is a paramount element to maintain and improve bone health throughout the life course [@Kohrt2004]. However, the mechanical loading osteogenic effect is highly dependent of the physical activity characteristics [@Burr2002]. Several animal studies [@Turner1998; @Turner2003; @Turner2005] helped to form our understanding of the principles that drive the bone tissue adaptative response. In general terms, high-impact dynamic loading, applied at high frequencies and including rest periods seems to favor increases in bone strength [@Turner2003]. This seems to corroborate with the available evidence from humans. For instance, findings from different types of sports have shown that athletes practicing modalities involving high- and/or odd-impact (e.g., volleyball and football) have a greater bone strength than compared to repetitive and low-impact sports (e.g., cycling and swimming) [@Nikander2010]. Nevertheless, the precise way in which the manipulation of mechanical loading variables influences the bone response in humans is not fully understand, particularly due to a difficulty in accurately assessing bone mechanical loading and its components.

Some questionnaires have been developed to capture bone-stimulating characteristics of specific activities [@Groothausen1997; @Weeks2008], but, although the scores generated by such subjective methods seem to be related to bone status, this association appears to be limited and inconsistent across studies [@Farr2011; @Kim2016; @Kim2018]. A more robust approach to assess mechanical loading is the use of equipment such as force plates, the gold standard for ground reaction forces measurement [@Medved2000]. However, the use of force plates is less applicable in some clinical and research settings, as it is an immobile and expensive device, limited to laboratory conditions. In the last few yers, objective methods based on wearable devices have been proposed as a solution to overcome these limitations, using acceleration data to estimate mechanical loading components, typically assessed only through force plates, namely peak ground reaction forces (pGRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020; @Veras2022] and peak loading rate (pLR) [@Fortune2014; @Veras2020; @Veras2022]. Although accelerometry-based equations have shown a high prediction accuracy [@Veras2020; @Veras2022], this versatile mechanical loading assessment approach is not widely used in either clinical practice or research yet. One of the issues that may contribute to this fact is the lack of validation studies covering a broader spectrum of activities, namely those known to have a greater osteogenic effect, such as high-impact activities.

Until now, studies that calibrate accelerometers to predict mechanical loading variables have usually employed only simple activities such as walking [@Fortune2014; @Neugebauer2018; @Veras2020] or walking and running [@Neugebauer2012; @Neugebauer2014; @Veras2022], while activities eliciting higher impact loads have not yet been tested. The fact that these accelerometry-based prediction equations are only valid for a limited set of activities hinders their ability to accurately quantify the amount of loading a person is submitted to daily, as the physical activities performed by a subject are often composed of other types of activity than just walking and running. Therefore, the validity of these methods should be expanded to other activities that are commonly used to induce positive bone adaptations. Jumping exercises, for instance, are frequently used as a strategy to improve bone health in post-menopausal women [@Du2021] and elder men [@Allison2015]. The direct measurement of the mechanical loading involved in these activities would greatly improve our ability to precisely control the loading regimen elicited by these physical exercises and to establish a more reliable relationship between mechanical stimulation and bone anabolic response and, consequently, to establish a clearer relationship between exercise dose and the extent of the bone adaptive response [@Hansford2022]. Therefore, the objective of this study was to develop accelerometry-based mechanical loading prediction equations for high-impact activities in adult subjects with a broad range of body masses, from normal weight to severe obesity.

# Methods {.unnumbered}

A total of `r max(sample_size$pGRF_N)` adults were enrolled in the study (`r unname(sample_size_sex["M"])` males; `r sample_descriptives$age_mean`±`r sample_descriptives$age_sd` years; `r sample_descriptives$height_mean`±`r sample_descriptives$height_sd` cm; `r sample_descriptives$body_mass_mean`±`r sample_descriptives$body_mass_sd` kg; `r sample_descriptives$BMI_mean`±`r sample_descriptives$BMI_sd` kg·m^-2^, X̄±SD). The protocol was approved by the local Ethics Committee (CES 192-14), and all participants were informed about the experiments purpose and protocol before giving written informed consent. No participant reported any musculoskeletal limitations. The measurement of height (stadiometer model 213; Seca, Hamburg, Germany) and body mass (digital scale model 899; Seca, Hamburg, Germany) was accomplished according to standard procedures.

In this study, jumps were chosen as the high-impact activity to be performed. After familiarization with the protocol, participants completed a series of trials involving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed starting from the floor up to 5cm, 15cm, and 30cm high boxes. Finally, continuous jumps were carried out at two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps, and both phases of the continuous jumps were recorded in a force plate (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions regarding arms movements, to land two-footed and to remain stationary on the force plate for five seconds after landing (except for the continuous jumps).

Throughout the protocol, three activity monitors (GT9X Link; 100Hz; ±16g range; ActiGraph, Pensacola, USA) were worn by each participant, at the following placements: i) at their right hip (along the anterior axillary line, at the level of the iliac crest); ii) at their lower back (at the midpoint between the two posterior superior iliac spines); and iii) at their right ankle (immediately superior to the lateral malleolus). Monitors worn at the hip and lower back were positioned on the same elastic belt with clips, and the ankle-worn monitor was fixed by an elastic belt and adhesive tape. Activity monitors were always positioned to ensure the alignment of the accelerometer vertical axis and the standing body longitudinal axis. The hip and lower back placements were chosen due to their proximity to the body center of mass, and the ankle placement was chosen as its location is closer to the ground contact point.

The activity monitors used include a primary and a secondary triaxial accelerometers. As the manufacturer proprietary filter is not applied on the secondary accelerometer raw data, only data from this accelerometer were used in this study, to improve data processing replicability. Data from force plates and accelerometers were operated through the manufacturer supplied software (for the force plates: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA) and exported as raw data from the x, y and z vectors. GRF was expressed in Newton (N) and acceleration in gravitational acceleration units (1*g* = 9.807 m·s^-2^).

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the ground reaction force signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both ground reaction force and acceleration signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the ground reaction force and acceleration resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak acceleration (pACC) was determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Then, pGRF was then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The pLR and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

\[f_i^{'} = \frac{f_{i + 1} - f_{i - 1}}{t_{i + 1} - t_{i - 1}}\]

\[f_{i_{max}}^{'} = max(f_i^{'})\]

Finally, pGRF, pACC, pLR and pAR means of the resultant vector and its vertical component for each participant at each jump type and height for each accelerometer placement were extracted and used in all remaining analyses.

Statistical analyses were performed through the R statistical software (version `r paste(R.Version()$major, R.Version()$minor, sep = ".")`, R Foundation for Statistical Computing, Vienna, Austria). All code used to generate the results can be found in an open platform (https://bit.ly/3szNSJS). Statistical significance was set as $\alpha$ = 0.05.

Linear mixed models were used to develop pGRF and pLR prediction equations for each vector (resultant and vertical) and accelerometer placement (ankle, lower back, and hip). Body mass was tested as a predictor in all models, along with the acceleration magnitude (pACC) in the pGRF models and its rate (pAR) in the pLR models. These predictors were entered as fixed effects and kept in the model, as all were shown to be significant (*p* < 0.05). Subject and the interaction between jump type and jump height were tested as random effects, and both factors were shown to improve the models. Final models were chosen according to the -2 log-likelihood statistic [@Field2012]. The conditional coefficient of determination (R^2^), which estimates the variance explained by both fixed and random factors [@Nakagawa2013], was computed.

Model validation was performed using the leave-one-out cross-validation [@Staudenmayer2012]. This method consists in separating data from one of the subjects into a testing dataset and keeping the remainder in the training dataset. Then, a new model is developed using the training dataset, with the same parameters that were defined using the entire sample. This new model is then applied to predict the outcome for the subject in the testing dataset. The whole process is repeated once for each participant (`r max(sample_size$pGRF_N)` times). Data from the testing dataset were used in all subsequent analyses.

The quality of the models was then evaluated both numerically and graphically. The numerical assessment was made by computing the following accuracy indices: the mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE). To visually analyze the agreement between the values predicted by the models and measured by the force plates, Bland-Altman plots were drawn. These plots are constructed by plotting the mean of the actual and predicted values on the x axis and their difference on the y axis. Then, the bias was calculated as the mean of these differences and plotted as a continuous horizontal line, and the limits of agreement were defined by bias ± 1.96 standard deviation and plotted as horizontal dashed lines [@Bland1986]. To further test if there were any systematic under- or over-estimation of the values by the models, one-sample *t*-tests were done to check whether the bias were significantly different from zero. Finally, linear regressions were employed to examine if the difference between actual and predicted values were affected by their magnitude, verifying, therefore, whether the prediction was constant throughout the entire magnitude range [@Giavarina2015].

# Results {.unnumbered}

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{fig1.png}
    \caption{Distribution of the resultant vector of mechanical loading (peak ground reaction force and peak loading rate) and acceleration (peak acceleration and peak acceleration rate) variables per jump type and height combination. Abbreviations: BW, body weight; pACC, peak acceleration; pAR, peak acceleration rate; pGRF, peak ground reaction force; pLR, peak loading rate.}
	\end{measuredfigure}
\end{figure}

In our protocol, three distinct jump types were performed, each one using several different heights. This variety of jump trials produced a considerable heterogeneity in the response of both the force plates- and accelerometer-measured variables. The results obtained for pGRF, pLR, pACC, and pAR, regarding their resultant vector, can be observed in Figure 1 (see Supplemental Figure S1 for the vertical vector). Panels A and B show the force plates-measured variables normalized by the subject’s body mass, the pGRF and pLR, respectively. It can be seen that, in the drop jump trials, there is a tendency for the loading magnitude to increase as the jump height increases. This tendency is also observed for the box jumps, but with a lower magnitude. As for the continuous jumps, both assayed heights elicited reasonably constant loadings. Overall, a similar pattern can be observed in the accelerometer-measured variables, pACC and pAR, in panels C and D, respectively. Nevertheless, this pattern is less clear, particularly for the continuous jumps. Regarding the influence of the accelerometer placement on the variables collected, in most of the trial’s, accelerometers worn at the hip and lower back produced approximately similar values, with a tendency for higher dispersion in data derived from the lower back accelerometers.

```{r accuracy_table, results = "show"}
source(here("tables", "tab1.R"))
tab1
```
Accelerometer-based mechanical loading prediction equations were developed based on linear mixed models. The regression coefficients, R^2^ and accuracy indices of the models derived from accelerometers worn at the ankle, lower back and hip, and from the resultant and vertical vectors are presented in Table 1. The R^2^ values for the pGRF models ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, showing that all the models were able to explain at least `r round(min(GRF_accuracy$R2_cond), 1) * 100`\% of the pGRF variance, and the observed MAPE values were between `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% and `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%. Despite similar accuracy results achieved by all equations, the model derived from data from the resultant vector of accelerometers worn at the hip were shown to have the lower accuracy error, represented by the smallest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%), and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).


Compared to the pGRF models, the pLR equations yielded poorer results. Although their R^2^ ranged from `r min(LR_accuracy$R2_cond)` to `r max(LR_accuracy$R2_cond)`, error measures were high, with MAPE ranging from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%. Nevertheless, similar to the pGRF models developed, the best pLR prediction equation in terms of prediction accuracy was the one from the resultant vector derived from hip-worn accelerometers, even thought error estimates were higher, with a MAE of `r myround(hip_LR_accuracy$MAE, 1)` N, MAPE of `r myround(hip_LR_accuracy$MAPE * 100, 1)`\%, and RMSE of `r myround(hip_LR_accuracy$RMSE, 1)` N.

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{fig2.png}
		\caption{Bland-Altman plots showing agreement between actual and predicted peak ground reaction force (Panel A) and loading rate (Panel B) for the resultant vector from hip-worn accelerometers. Continuous thick lines show bias (average of the differences between actual and predicted values) while dashed lines show the limits of agreement (bias \pm 1.96 standard deviation). Abbreviations: pGRF, peak ground reaction force; pLR, peak loading rate.}
	\end{measuredfigure}
\end{figure}

To visualize the agreement between the pGRF and pLR values measured by the force plates and predicted by the models, a series of Bland-Altman plots were built for both the resultant and vertical vectors and for all accelerometer placements. Figure 2 shows the Bland-Altman plots for the pGRF (panel A) and pLR (panel B) resultant vector of hip-worn accelerometers. Both plots have shown good agreement levels, as at least `r ba_agreement`\% of the data points are within the limits of agreement and the data tended to cluster around zero. Bland-Altman plots for the other accelerometer placements and vectors have also shown the same previously described patterns and can be found in Supplemental Figure S2. Also, as could be observed by the one-sample *t*-tests results, none of the equations presented a bias significantly different than zero (*p* > 0.05). Furthermore, all of the equations have shown a proportional bias, i.e., the magnitude of the values influences the prediction error. However, as the linear regressions used to assess the proportional bias presented a very small R^2^ (maximum value of `r myround(max(ba_regression_ankle_R2), 2)`), this effect is neglectable.

Finally, an R package was also developed [@Veras2021] to simplify the application of all the necessary data processing steps involved in the implementation of the developed prediction equations. The package website (https://verasls.github.io/impactr) has all the installation instructions and documentation necessary to apply this method on the user’s own data.

# Discussion {.unnumbered}

This study aimed to develop accelerometer-based mechanical loading prediction equations, namely for pGRF and pLR, for jumping exercises of different types and heights, as well as to test different accelerometer placements and data derived from both the resultant and vertical acceleration vectors. The results have shown that pGRF prediction models for jumping activities based on accelerometry data have a MAPE around 13%, while pLR prediction equations resulted in a lower accuracy, with a MAPE around 22%. Also, these models are the first to be developed specifically for high-impact jumping activities in adults.

Bone is known to increase its tissue mass and enhance its micro-architecture arrangement in response to mechanical stimulation [@Turner2003; @Fonseca2014], particularly of higher magnitudes [@Vainionpaa2006; @Stiles2017]. These types of high-intensity impact activities have shown to be more associated with bone health improvements, even if they are usually performed in low volumes throughout the day. In fact, Vainionpää et al [@Vainionpaa2006] demonstrated that performing a relatively low number (<100) of pACC above 4.9g daily is positively associated with increases in bone mineral density. Usually, this acceleration magnitude can be achieved mostly by specific movements such as jumps [@Vainionpaa2006; @Rowlands2012] . Nevertheless, accelerometry-based mechanical loading regression models have mostly been validated for a set of activities, such as walking and running, that elicit relatively low acceleration magnitudes [@Neugebauer2014; @Veras2020; @Veras2022]. The results of this study provide new mechanical loading prediction models for jumping, which are especially important as it is one of the most popular and effective activities prescribed for bone health improvement [@Beck2017].

The development of such models was done with the direct purpose of allowing to assess the loading sustained by the skeletal system during daily physical activity. Nowadays, physical activity monitorization and exercise prescription aiming bone health promotion has been carried out almost blindly or through the use of subjective instruments such as questionaries. Moreover, the absence of a trustworthy method to monitor mechanical loading on a daily basis may preclude the prescription and self-involvement of patients in high-impact activities that may expose them to a higher risk of joint injuries. These issues may be overcome with the recent popularization of wearable sensors, such as accelerometers, and their ability to be used as a tool for mechanical loading prediction. This would allow some important clinical applications, as the models developed in this study, when used in combination with other prediction equations for other types of activities [@Veras2022], can be used to fully analyze the mechanical loading spectrum that people are submitted to daily and relate to bone health parameters. Therefore, a more complete understanding can be achieved on how mechanical loading influences bone health, by, for instance, identifying the loading volume and magnitude range that are more efficient in inducing bone health benefits. Also, the development and distribution of an R package that automates the steps necessary to apply these models enhances their applicability to potential users [@Veras2021]. This can prove to be an important tool, as the application of these models depend on some laborious steps, such as the application of digital filters to reduce noise from the signal, detection of the peaks in this signal and the computation of derivatives to obtain the acceleration rate, necessary for the pRL prediction.

When comparing the accuracy indices of the pGRF prediction equations developed in the current study with other equations from past investigations, developed for walking and running, it can be seen that our models have a lower accuracy [@Neugebauer2012; @Neugebauer2014; @Neugebauer2018; @Veras2020; @Veras2022]. This can, at least partially, be explained by the high heterogeneity in the jumping movement patterns, as several combinations of jump type and height were tested, and also that subjects with different skill levels were included in the sample. Nevertheless, these MAPE values are similar to the results usually found in accelerometer-based energy expenditure prediction [@Diniz-Sousa2020], which have widespread use in research. As for the pLR prediction equations, they presented a lower accuracy compared to the pGRF equations, which corroborates with previous findings [@Fortune2014; @Veras2020; @Veras2022]. This is possibly due to the fact that loading rate is typically calculated through a derivative of the force signal from the beginning of the foot contact to the floor to the point where the pGRF is achieved [@Liikavainio2007; @Fortune2014; @Veras2020; @Veras2022]. As it is a small time window, usually less than 0.2 seconds for the jumps that were performed in our protocol, it requires a fine sampling acquisition that captures small graduations on the force values throughout time, which is difficult to achieve with common accelerometer sampling frequencies, that are usually 100Hz at maximum. Nevertheless, although our pLR prediction models had a worse accuracy compared with the pGRF models, its accuracy was similar or even better than the results from previous studies that developed models to predict pLR during walking only [@Fortune2014; @Veras2020].

In our protocol, three accelerometer placements (ankle, lower back, and hip) and two vectors (resultant and vertical) have been tested. Overall, the prediction models performance was reasonably consistent among the different placement and vector combinations, allowing some conclusions can be drawn. First, the hip accelerometer placement which, traditionally, has been widely employed in research for the direct measurement of physical activity [@Migueles2017], yielded a slightly better accuracy compared to the other placements tested. This could be useful, as previous data collected from hip-worn accelerometers, even for other purposes, can be adequately used to obtain mechanical loading predictions. Also, studies that aim to extract mechanical loading together with other variables derived from accelerometry data and that use the hip as the default accelerometer placement, can reliably use this placement. Second, as the magnitude of the resultant vector of both pGRF and pLR, in the activities that were tested, is derived mostly from the vertical vector, and there were no considerable differences in performance between models using both vectors, they can be used interchangeably to predict mechanical loading. Besides that, while models using data from the vertical vector allow the use of uniaxial accelerometers, data derived from the resultant vector have the important advantage of not depending on the correct accelerometer axes orientation, which cannot be guaranteed during data collection in free-living conditions [@Welk2005].

Although the models presented in this study showed to be valid, some limitations ought to be considered. First, the choice of activity to elicit high-impact loading was jumping, and several combinations of jump type and height were tested. Nevertheless, the results presented here cannot be assured for other high-impact activities, or even for other types of jumps that were not tested. Second, no external sample for model validation was recruited. However, the leave-one-out cross-validation method that was used is the recommended strategy in these situations [@Staudenmayer2012]. Third, to apply these models in accelerometer-recorded free-living physical activity data, there is a need to identify, in the acceleration signal, the periods in which the activity being performed is jumping. Therefore, methods for physical activity pattern detection should be used in combination with these models.

In conclusion, accelerometry-based mechanical loading prediction models, especially for pGRF, have proved to be a valid method for determining the mechanical stimulation induced by jumps of various types and heights. These prediction models were validated using data from both the resultant and vertical vectors and with accelerometers placed at the ankle, lower back and hip. Results from this study will enable the continuous assessment of mechanical loading in clinical settings, providing means to objectively determine the osteogenic potential of daily physical activity and to better monitor and prescribe exercise aimed to improve bone health. Compared to pGRF, prediction of pLR showed lower accuracy.

# Acknowledgements {.unnumbered}

This study was funded by the Foundation for Science and Technology (FCT) [grant number PTDC/DTP-DES/0968/2014] and by the European Regional Development Fund (ERDF) through the Operational Competitiveness Programme (COMPETE) [grant number POCI-01-0145-FEDER-016707]. The study was developed in the Research Centre in Physical Activity, Health and Leisure (CIAFEL) funded by ERDF through the COMPETE and by the FCT [grant number UID/DTP/00617/2020], and Laboratory for Integrative and Translational Research in Population Health (ITR), funded by the FCT [grant number LA/P/0064/2020].  Lucas  Veras and  Giorjines Boppre are supported by the FCT [grant numbers UI/BD/150673/2020 and SFRH/BD/146976/2019]. The authors thank the participants who took part in this research and all who have collaborated in the research project.

\newpage

# References {.unnumbered}

<div id="refs"></div>

\newpage

# Supplemental Material {.unnumbered}

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=0.7\textwidth]{figS1.png}
	\end{measuredfigure}
\end{figure}
\footnotesize \textbf{Supplemental Figure S1.} An example of a participant performing each of the jump types assayed. Panels A and B show the initial (take-off phase) and final (landing phase) phases of the drop jumps, respectively, while Panels C and D show the initial and final phases of the box jumps and Panels E and F show the initial and final phase of the continuous jumps.

\newpage

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=0.7\textwidth]{figS2.png}
	\end{measuredfigure}
\end{figure}
\footnotesize \textbf{Supplemental Figure S2.} Accelerometers positioning on the body. Details of the accelerometer positioning on the ankle, lower back and hip are depicted in Panels A, B and C, respectively. Panel D shows all accelerometer placements simultaneously.

\newpage

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{figS3.png}
	\end{measuredfigure}
\end{figure}
\footnotesize \textbf{Supplemental Figure S3.} Distribution of the vertical vector of mechanical loading (peak ground reaction force and peak loading rate) and acceleration (peak acceleration and peak acceleration rate) variables per jump type and height combination. Abbreviations: BW, body weight; pACC, peak acceleration; pAR, peak acceleration rate; pGRF, peak ground reaction force; pLR, peak loading rate.

\newpage

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{figS4.png}
	\end{measuredfigure}
\end{figure}
\footnotesize \textbf{Supplemental Figure S4.} Bland-Altman plots showing agreement between actual and predicted peak ground reaction force for accelerometers worn at ankle (Panels A and D), lower back (Panels B and E) and hip (Panels C and F), for both the resultant vector (Panels A to C) and its vertical component (Panels D to F). Continuous thick lines show bias (average of the differences between actual and predicted values) while dashed lines show the limits of agreement (bias ± 1.96 standard deviation). Abbreviations: pGRF, peak ground reaction force.

\newpage

\begin{figure}[ht]
	\captionsetup{singlelinecheck = false, format = hang, justification = raggedright, font = footnotesize, labelsep = space}
	\centering
	\begin{measuredfigure}
		\includegraphics[width=\textwidth]{figS5.png}
	\end{measuredfigure}
\end{figure}
\footnotesize \textbf{Supplemental Figure S5.} Bland-Altman plots showing agreement between actual and predicted peak loading rate for accelerometers worn at ankle (Panels A and D), lower back (Panels B and E) and hip (Panels C and F), for both the resultant vector (Panels A to C) and its vertical component (Panels D to F). Continuous thick lines show bias (average of the differences between actual and predicted values) while dashed lines show the limits of agreement (bias ± 1.96 standard deviation). Abbreviations: pLR, peak loading rate.
